---
title: "Sub-Saharan Africa Diabetes Burden Analysis"
subtitle: "GBD 2023 regional and subregional trends, forecasts, and burden composition"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
geometry: margin=0.75in
fontsize: 9pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{array}
---

```{r setup, include=FALSE}
# Resolve the project root so the report can be rendered from the reports folder.
project_root <- normalizePath("..")

# Make all relative paths resolve from the project root during rendering.
knitr::opts_knit$set(root.dir = project_root)

# Hide code in the PDF while keeping source chunks commented and readable.
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.pos = "H")

# Load only the packages required directly in this report.
suppressPackageStartupMessages({
  library(dplyr)
  library(forecast)
  library(ggplot2)
  library(knitr)
  library(readr)
  library(scales)
  library(tidyr)
})

# Shared cleaning, validation, forecasting, and plotting helpers.
source(file.path(project_root, "scripts/common/report_helpers.R"))
```

```{r paths-and-constants, include=FALSE}

# Create output directories for machine-readable tables and publication figures.
output_dir <- file.path(project_root, "outputs/sub_saharan_africa/full_analysis")
table_dir <- file.path(output_dir, "tables")
figure_dir <- file.path(output_dir, "figures")
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

# Store all input paths in one object so the availability table and readers agree.
paths <- list(
  rates = file.path(project_root, "data/raw/gbd_2023/sub_saharan_africa/ssa_and_subregions_age_standardized_rates_1990_2023.csv"),
  counts = file.path(project_root, "data/raw/gbd_2023/sub_saharan_africa/ssa_and_subregions_all_age_counts_1990_2023.csv"),
  age_counts = file.path(project_root, "data/raw/gbd_2023/sub_saharan_africa/ssa_and_subregions_age_specific_counts_1990_2023.csv"),
  age_rates = file.path(project_root, "data/raw/gbd_2023/sub_saharan_africa/ssa_and_subregions_age_specific_rates_1990_2023.csv"),
  population = file.path(project_root, "data/raw/un_wpp_2024/WPP2024_PopulationByAge5GroupSex_Medium.csv.gz")
)

# Fixed region order used in tables, legends, and facets.
locations <- c(
  "Sub-Saharan Africa",
  "Central Sub-Saharan Africa",
  "Eastern Sub-Saharan Africa",
  "Southern Sub-Saharan Africa",
  "Western Sub-Saharan Africa"
)

# Fixed cause and sex orders keep outputs stable between renders.
cause_levels <- c("Diabetes mellitus", "Diabetes mellitus type 1", "Diabetes mellitus type 2")
sex_levels <- c("Male", "Female", "Both")

# Convert long GBD measure names to compact plotting labels.
measure_short_df <- tibble(
  measure_name = names(measure_short),
  measure = unname(measure_short)
)

# Use an accessible, multi-hue palette for regional comparisons.
region_palette <- c(
  "Sub-Saharan Africa" = "#111111",
  "Central Sub-Saharan Africa" = "#0072B2",
  "Eastern Sub-Saharan Africa" = "#009E73",
  "Southern Sub-Saharan Africa" = "#D55E00",
  "Western Sub-Saharan Africa" = "#CC79A7"
)

# A compact paper theme shared by all ggplot figures in the report.
paper_theme <- function(base_size = 8) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(face = "bold", size = base_size),
      axis.title = element_text(size = base_size),
      axis.text = element_text(size = base_size - 1),
      legend.position = "bottom",
      legend.title = element_blank(),
      plot.caption = element_text(size = base_size - 1)
    )
}

# Preserve decimals for small rates such as T1DM incidence and deaths.
rate_axis_labels <- function(x) {
  ifelse(abs(x) < 10, sprintf("%.1f", x), comma(x, accuracy = 1))
}
```


```{r helper-functions, include=FALSE}
read_stage2 <- function(path, expected_age, expected_metric) {
  # Missing optional files return NULL so pending sections can remain in place.
  if (!file.exists(path)) return(NULL)

  # Read the requested GBD extract and retain the 1990-2023 analysis window.
  data <- read_csv(path, show_col_types = FALSE) %>%
    filter(year %in% 1990:2023)

  # Fail early if the extract is not shaped like a GBD stage-2 file.
  missing_columns <- setdiff(required_gbd_columns, names(data))
  if (length(missing_columns) > 0) {
    stop("Missing required columns in ", path, ": ", paste(missing_columns, collapse = ", "))
  }

  # Validate dimensions, year coverage, uniqueness, missingness, and UI bounds.
  key <- c("measure_id", "location_id", "sex_id", "age_id", "cause_id", "metric_id", "year")
  checks <- c(
    expected_locations = setequal(data$location_name, locations),
    expected_age = setequal(data$age_name, expected_age),
    expected_metric = identical(unique(data$metric_name), expected_metric),
    complete_years = setequal(data$year, 1990:2023),
    expected_sexes = setequal(data$sex_name, sex_levels),
    expected_causes = setequal(data$cause_name, cause_levels),
    no_duplicate_keys = !anyDuplicated(data[key]),
    no_missing_values = !anyNA(data[required_gbd_columns]),
    valid_bounds = all(data$lower <= data$val & data$val <= data$upper)
  )
  if (!all(checks)) {
    stop("Input validation failed for ", path, ": ", paste(names(checks)[!checks], collapse = ", "))
  }

  # Return a validated analysis-ready data frame.
  data
}
```

```{r read-inputs, include=FALSE}
# Age-standardized rates drive trend, comparison, and forecast analyses.
rates <- read_stage2(paths$rates, "Age-standardized", "Rate")

# All-age counts support absolute-burden and composition analyses.
counts <- read_stage2(paths$counts, "All ages", "Number")

# Age-specific counts are available for quality checks and future decomposition.
age_counts <- read_stage2(paths$age_counts, c(
  "0-6 days", "7-27 days", "1-5 months", "6-11 months", "12-23 months",
  "2-4 years", "5-9 years", "10-14 years", "15-19 years", "20-24 years",
  "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years",
  "50-54 years", "55-59 years", "60-64 years", "65-69 years", "70-74 years",
  "75-79 years", "80-84 years", "85-89 years", "90-94 years", "95+ years"
), "Number")

# Age-specific rates are required for the completed Stage 2 analysis.
age_rates <- read_stage2(paths$age_rates, c(
  "0-6 days", "7-27 days", "1-5 months", "6-11 months", "12-23 months",
  "2-4 years", "5-9 years", "10-14 years", "15-19 years", "20-24 years",
  "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years",
  "50-54 years", "55-59 years", "60-64 years", "65-69 years", "70-74 years",
  "75-79 years", "80-84 years", "85-89 years", "90-94 years", "95+ years"
), "Rate")

# All four GBD inputs are required for the completed report to render safely.
if (is.null(rates) || is.null(counts) || is.null(age_counts) || is.null(age_rates)) {
  stop("This report requires age-standardized rates, all-age counts, age-specific counts, and age-specific rates.")
}
```

```{r input-availability, include=FALSE}
# Summarise which analyses are currently enabled by the available files.
availability <- tibble(
  input = c(
    "Age-standardized rates", "All-age counts", "Age-specific counts",
    "Age-specific rates", "UN WPP 2024 population"
  ),
  file = unlist(paths, use.names = FALSE),
  status = c(
    "AVAILABLE",
    "AVAILABLE",
    "AVAILABLE",
    "AVAILABLE",
    if (file.exists(paths$population)) "AVAILABLE" else "MISSING"
  ),
  enabled_analysis = c(
    "Endpoints, trends, sex ratios, forecasts",
    "Count endpoints, diabetes-type shares, YLL-to-YLD ratios",
    "Age-specific count validation and readiness for decomposition",
    "GBD-consistent decomposition, rate-effect estimation, age-period-cohort analysis",
    "Demographic-only projections and population sensitivity analysis"
  )
)

lstm_status <- lstm_backend_status()

# Register all forecast model families and whether they are active in this render.
model_registry <- tibble(
  model = c("Log-linear trend", "ARIMA", "ETS", "LSTM"),
  status = c(
    "Implemented",
    "Implemented",
    "Implemented",
    if (lstm_status$lstm_available) "Implemented and fitted for headline comparison" else "Specified; backend pending"
  ),
  primary_use = c(
    "Average annual percentage change over 1990-2023",
    "Short-term age-standardized rate forecasting",
    "Short-term age-standardized rate forecasting",
    "Machine-learning forecast candidate for nonlinear temporal patterns"
  )
)
```

```{r summary-helper-functions, include=FALSE}
endpoint_by_location <- function(data) {
  # Produce 1990 and 2023 endpoints separately for each location.
  data %>%
    group_by(location_name) %>%
    group_modify(~ endpoint_summary(.x)) %>%
    ungroup()
}

trend_by_location <- function(data) {
  # Estimate log-linear APC trends separately for each location.
  data %>%
    group_by(location_name) %>%
    group_modify(~ log_linear_trends(.x)) %>%
    ungroup()
}

forecast_by_location <- function(data) {
  # Run ARIMA/ETS expanding-window validation and selected-model forecasts.
  data %>%
    group_by(location_name) %>%
    group_map(~ {
      result <- generate_rate_forecasts(.x)
      list(
        forecasts = result$forecasts %>% mutate(location_name = .y$location_name),
        accuracy = result$accuracy %>% mutate(location_name = .y$location_name)
      )
    })
}

# Average marginal contributions over all six factor orderings. This is the
# symmetric (Shapley) decomposition for population size, age structure, and
# age-specific rate.
decompose_three_factors <- function(data) {
  p <- data %>% distinct(year, age_id, population) %>%
    arrange(year, age_id) %>%
    group_by(year) %>%
    mutate(total_population = sum(population), age_share = population / total_population) %>%
    ungroup()
  r <- data %>% select(year, age_id, rate = val) %>% arrange(year, age_id)
  x <- left_join(p, r, by = c("year", "age_id"))
  n <- setNames(x %>% distinct(year, total_population) %>% pull(total_population),
                x %>% distinct(year, total_population) %>% pull(year))
  shares <- split(x$age_share, x$year)
  rates_by_year <- split(x$rate, x$year)
  value <- function(state) {
    size_year <- if (state[["size"]]) "2023" else "1990"
    structure_year <- if (state[["structure"]]) "2023" else "1990"
    rate_year <- if (state[["rate"]]) "2023" else "1990"
    n[[size_year]] * sum(shares[[structure_year]] * rates_by_year[[rate_year]]) / 1e5
  }
  permutations <- list(
    c("size", "structure", "rate"), c("size", "rate", "structure"),
    c("structure", "size", "rate"), c("structure", "rate", "size"),
    c("rate", "size", "structure"), c("rate", "structure", "size")
  )
  contributions <- lapply(permutations, function(ordering) {
    state <- c(size = FALSE, structure = FALSE, rate = FALSE)
    out <- c(size = 0, structure = 0, rate = 0)
    previous <- value(state)
    for (factor_name in ordering) {
      state[[factor_name]] <- TRUE
      current <- value(state)
      out[[factor_name]] <- current - previous
      previous <- current
    }
    out
  })
  effects <- rowMeans(do.call(cbind, contributions))
  baseline <- value(c(size = FALSE, structure = FALSE, rate = FALSE))
  endpoint <- value(c(size = TRUE, structure = TRUE, rate = TRUE))
  tibble(
    count_1990_reconstructed = baseline,
    count_2023_reconstructed = endpoint,
    observed_change = endpoint - baseline,
    population_size_effect = effects[["size"]],
    age_structure_effect = effects[["structure"]],
    age_specific_rate_effect = effects[["rate"]],
    decomposition_error = sum(effects) - (endpoint - baseline)
  )
}
```



```{r primary-summaries, include=FALSE}
# Endpoint tables describe baseline, latest value, and percentage change.
rate_endpoints <- endpoint_by_location(rates)
count_endpoints <- endpoint_by_location(counts)

# Trend models provide interpretable APC estimates for every stratum.
rate_trends <- trend_by_location(rates)

# Forecast helpers return both projected values and validation metrics.
rate_forecast_parts <- forecast_by_location(rates)
rate_forecasts <- bind_rows(lapply(rate_forecast_parts, `[[`, "forecasts")) %>%
  relocate(location_name)
forecast_accuracy <- bind_rows(lapply(rate_forecast_parts, `[[`, "accuracy")) %>%
  relocate(location_name)

# Run the fitted headline ARIMA/ETS/LSTM comparison.
headline_model_comparison <- compare_headline_forecast_models(rates)
headline_model_accuracy <- headline_model_comparison$accuracy
headline_model_errors <- headline_model_comparison$errors
headline_selected_forecasts <- generate_headline_selected_forecasts(
  rates,
  headline_model_accuracy,
  headline_model_errors
)
```

```{r secondary-summaries, include=FALSE}
# Compare male and female age-standardized rates by year and measure.
sex_ratios <- rates %>%
  group_by(location_name) %>%
  group_modify(~ sex_rate_ratios(.x)) %>%
  ungroup()

# Estimate the contribution of T1DM and T2DM to all-age diabetes burden.
type_shares <- counts %>%
  group_by(location_name) %>%
  group_modify(~ diabetes_type_shares(.x)) %>%
  ungroup()

# YLL:YLD ratios show whether fatal or non-fatal burden dominates.
yll_yld_ratios <- counts %>%
  group_by(location_name) %>%
  group_modify(~ yll_yld_ratio(.x)) %>%
  ungroup()

# Smooth YLL:YLD ratios into interpretable calendar periods.
period_ratios <- yll_yld_ratios %>%
  mutate(
    period = case_when(
      year <= 1994 ~ "1990-1994",
      year <= 1999 ~ "1995-1999",
      year <= 2004 ~ "2000-2004",
      year <= 2009 ~ "2005-2009",
      year <= 2014 ~ "2010-2014",
      year <= 2019 ~ "2015-2019",
      TRUE ~ "2020-2023"
    )
  ) %>%
  group_by(location_name, period) %>%
  summarise(yll_to_yld_ratio = mean(yll_to_yld_ratio), .groups = "drop")

# Confirm that age-specific count strata contain the expected dimensions.
age_count_qc <- age_counts %>%
  group_by(location_name, sex_name) %>%
  summarise(
    rows = n(),
    measures = n_distinct(measure_name),
    causes = n_distinct(cause_name),
    ages = n_distinct(age_name),
    years = n_distinct(year),
    .groups = "drop"
  )

# Match counts and rates to reconstruct GBD-implied demographic denominators.
age_population_candidates <- age_counts %>%
  select(location_id, location_name, sex_id, sex_name, age_id, age_name,
         cause_id, measure_id, year, count = val) %>%
  inner_join(
    age_rates %>% select(location_id, sex_id, age_id, cause_id, measure_id,
                         year, rate = val),
    by = c("location_id", "sex_id", "age_id", "cause_id", "measure_id", "year")
  ) %>%
  filter(rate > 0) %>%
  mutate(implied_population = count / rate * 1e5)

age_population <- age_population_candidates %>%
  group_by(location_id, location_name, sex_id, sex_name, age_id, age_name, year) %>%
  summarise(
    population = median(implied_population),
    denominator_estimates = n(),
    max_relative_spread = if_else(
      population > 0,
      max(abs(implied_population - population) / population),
      0
    ),
    .groups = "drop"
  )

population_qc <- age_population %>%
  summarise(
    demographic_cells = n(),
    cells_with_denominator = sum(denominator_estimates > 0),
    maximum_relative_spread = max(max_relative_spread),
    median_relative_spread = median(max_relative_spread)
  )

# Symmetric decomposition for every region, sex, cause, and measure.
decomposition_data <- age_rates %>%
  filter(year %in% c(1990, 2023)) %>%
  inner_join(
    age_population %>% select(location_id, sex_id, age_id, year, population),
    by = c("location_id", "sex_id", "age_id", "year")
  )

symmetric_decomposition <- decomposition_data %>%
  group_by(location_name, sex_name, cause_name, measure_name) %>%
  group_modify(~ decompose_three_factors(.x)) %>%
  ungroup() %>%
  mutate(
    population_size_pct = 100 * population_size_effect / observed_change,
    age_structure_pct = 100 * age_structure_effect / observed_change,
    age_specific_rate_pct = 100 * age_specific_rate_effect / observed_change
  )

# Age-specific endpoints and log-linear changes provide directly interpretable
# age patterns without imposing an unidentified full APC parameterization.
age_specific_endpoints <- age_rates %>%
  filter(year %in% c(1990, 2023)) %>%
  select(location_name, sex_name, cause_name, measure_name, age_id, age_name,
         year, val) %>%
  pivot_wider(names_from = year, values_from = val, names_prefix = "rate_") %>%
  mutate(
    absolute_change = rate_2023 - rate_1990,
    relative_change_pct = if_else(rate_1990 > 0, 100 * (rate_2023 / rate_1990 - 1), NA_real_)
  )

# Estimate age-adjusted net drift for incidence and mortality. This is a
# constrained age-period trend model, not a full identifiable APC model.
age_adjusted_drift <- age_counts %>%
  filter(measure_name %in% c("Incidence", "Deaths")) %>%
  inner_join(
    age_population %>% select(location_id, sex_id, age_id, year, population),
    by = c("location_id", "sex_id", "age_id", "year")
  ) %>%
  filter(population > 0, val >= 0) %>%
  group_by(location_name, sex_name, cause_name, measure_name) %>%
  group_modify(~ {
    fit <- glm(val ~ factor(age_id) + year, offset = log(population),
               family = quasipoisson(link = "log"), data = .x)
    coefficient <- coef(summary(fit))["year", ]
    tibble(
      annual_drift_pct = 100 * (exp(coefficient[["Estimate"]]) - 1),
      lower_95 = 100 * (exp(coefficient[["Estimate"]] - 1.96 * coefficient[["Std. Error"]]) - 1),
      upper_95 = 100 * (exp(coefficient[["Estimate"]] + 1.96 * coefficient[["Std. Error"]]) - 1),
      p_value = coefficient[["Pr(>|t|)"]]
    )
  }) %>%
  ungroup()
```



```{r forecast-diagnostics, include=FALSE}
# Restrict diagnostics to the headline overall SSA, both-sexes, total diabetes series.
headline_series <- rates %>%
  filter(
    location_name == "Sub-Saharan Africa",
    cause_name == "Diabetes mellitus",
    sex_name == "Both"
  ) %>%
  arrange(factor(measure_name, levels = measure_levels), year)

# Calculate trailing means to show medium-term direction without removing raw series.
trailing_mean_data <- headline_series %>%
  group_by(measure_name) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(trailing_mean_5yr = zoo::rollmean(val, k = 5, fill = NA, align = "right")) %>%
  ungroup()

# Compare differencing suggestions and the order chosen by auto.arima().
stationarity_checks <- headline_series %>%
  group_by(measure_name) %>%
  group_modify(~ {
    y <- ts(.x$val, start = min(.x$year), frequency = 1)
    fit <- auto.arima(y)
    tibble(
      kpss_ndiffs = ndiffs(y, test = "kpss"),
      adf_ndiffs = ndiffs(y, test = "adf"),
      pp_ndiffs = ndiffs(y, test = "pp"),
      auto_arima_p = arimaorder(fit)[["p"]],
      auto_arima_d = arimaorder(fit)[["d"]],
      auto_arima_q = arimaorder(fit)[["q"]],
      selected_arima = paste0(
        "ARIMA(",
        arimaorder(fit)[["p"]], ",",
        arimaorder(fit)[["d"]], ",",
        arimaorder(fit)[["q"]], ")"
      )
    )
  }) %>%
  ungroup() %>%
  mutate(Measure = recode(measure_name, !!!measure_short)) %>%
  select(Measure, everything(), -measure_name)

# Extract ACF/PACF values in data-frame form for a consistent ggplot layout.
acf_pacf_data <- headline_series %>%
  group_by(measure_name) %>%
  group_modify(~ {
    y <- ts(.x$val, start = min(.x$year), frequency = 1)
    acf_values <- acf(y, plot = FALSE, lag.max = 12)
    pacf_values <- pacf(y, plot = FALSE, lag.max = 12)
    bind_rows(
      tibble(
        lag = as.integer(acf_values$lag[-1]),
        value = as.numeric(acf_values$acf[-1]),
        diagnostic = "ACF"
      ),
      tibble(
        lag = as.integer(pacf_values$lag),
        value = as.numeric(pacf_values$acf),
        diagnostic = "PACF"
      )
    )
  }) %>%
  ungroup() %>%
  left_join(measure_short_df, by = "measure_name") %>%
  mutate(measure = factor(measure, levels = unname(measure_short)))
```

```{r write-derived-csvs, include=FALSE}
# Write every derived table needed for reproducibility outside the PDF.
write_csv(availability, file.path(table_dir, "input_availability.csv"))
write_csv(rate_endpoints, file.path(table_dir, "age_standardized_rate_endpoints.csv"))
write_csv(rate_trends, file.path(table_dir, "age_standardized_log_linear_trends.csv"))
write_csv(rate_forecasts, file.path(table_dir, "age_standardized_rate_forecasts_2024_2030.csv"))
write_csv(forecast_accuracy, file.path(table_dir, "forecast_model_accuracy.csv"))
write_csv(count_endpoints, file.path(table_dir, "all_age_count_endpoints.csv"))
write_csv(sex_ratios, file.path(table_dir, "sex_rate_ratios.csv"))
write_csv(type_shares, file.path(table_dir, "diabetes_type_shares.csv"))
write_csv(yll_yld_ratios, file.path(table_dir, "yll_to_yld_ratio_annual.csv"))
write_csv(period_ratios, file.path(table_dir, "yll_to_yld_ratio_periods.csv"))
write_csv(age_count_qc, file.path(table_dir, "age_specific_count_quality_summary.csv"))
write_csv(model_registry, file.path(table_dir, "forecast_model_registry.csv"))
write_csv(stationarity_checks, file.path(table_dir, "forecast_stationarity_checks.csv"))
write_csv(acf_pacf_data, file.path(table_dir, "forecast_acf_pacf_values.csv"))
write_csv(headline_model_accuracy, file.path(table_dir, "headline_forecast_model_comparison.csv"))
write_csv(headline_model_errors, file.path(table_dir, "headline_forecast_model_validation_predictions.csv"))
write_csv(headline_selected_forecasts, file.path(table_dir, "headline_selected_model_forecasts_2024_2030.csv"))
write_csv(age_population, file.path(table_dir, "gbd_implied_age_sex_population_1990_2023.csv"))
write_csv(population_qc, file.path(table_dir, "gbd_implied_population_validation.csv"))
write_csv(symmetric_decomposition, file.path(table_dir, "symmetric_decomposition_1990_2023.csv"))
write_csv(age_specific_endpoints, file.path(table_dir, "age_specific_rate_endpoints_1990_2023.csv"))
write_csv(age_adjusted_drift, file.path(table_dir, "age_adjusted_incidence_mortality_drift.csv"))
```

```{r table-formatting-helpers, include=FALSE}
clean_measure <- function(x) recode(x, !!!measure_short)
clean_location <- function(x) recode(
  x,
  "Sub-Saharan Africa" = "SSA",
  "Central Sub-Saharan Africa" = "Central SSA",
  "Eastern Sub-Saharan Africa" = "Eastern SSA",
  "Southern Sub-Saharan Africa" = "Southern SSA",
  "Western Sub-Saharan Africa" = "Western SSA"
)
clean_cause <- function(x) recode(
  x,
  "Diabetes mellitus" = "DM",
  "Diabetes mellitus type 1" = "T1DM",
  "Diabetes mellitus type 2" = "T2DM"
)
fmt_ci <- function(x, lo, hi, digits = 1) {
  sprintf(
    paste0("%.", digits, "f (%.", digits, "f-%.", digits, "f)"),
    x, lo, hi
  )
}
fmt_p <- function(x) format.pval(x, digits = 2, eps = 0.001)
format_notes <- function(notes) {
  if (is.null(notes) || length(notes) == 0) return("")
  notes <- gsub("%", "\\\\%", notes, fixed = TRUE)
  paste0(
    "\n\\vspace{1mm}\n\\begin{minipage}{0.96\\linewidth}\n",
    "\\scriptsize\\textit{Notes:} ",
    paste(notes, collapse = " "),
    "\n\\end{minipage}"
  )
}
format_pdf_table <- function(
    data, caption, digits = 1, max_rows = 40, font_size = "scriptsize",
    notes = NULL) {
  shown <- head(data, max_rows)
  numeric_columns <- vapply(shown, is.numeric, logical(1))
  shown[numeric_columns] <- lapply(shown[numeric_columns], round, digits = digits)
  paste0(
    "\\begingroup\n\\", font_size, "\n",
    "\\setlength{\\tabcolsep}{2.5pt}\n",
    "\\renewcommand{\\arraystretch}{0.9}\n",
    kable(
      shown, format = "latex", booktabs = TRUE, longtable = TRUE,
      caption = caption, escape = TRUE
    ),
    format_notes(notes),
    "\n\\endgroup"
  )
}
```

```{r key-finding-values, include=FALSE}
overall_prevalence <- rate_endpoints %>%
  filter(
    location_name == "Sub-Saharan Africa",
    cause_name == "Diabetes mellitus",
    sex_name == "Both",
    measure_name == "Prevalence"
  )
overall_deaths <- rate_endpoints %>%
  filter(
    location_name == "Sub-Saharan Africa",
    cause_name == "Diabetes mellitus",
    sex_name == "Both",
    measure_name == "Deaths"
  )
highest_2023_prevalence <- rate_endpoints %>%
  filter(cause_name == "Diabetes mellitus", sex_name == "Both", measure_name == "Prevalence") %>%
  slice_max(value_2023, n = 1, with_ties = FALSE)
fastest_daly_growth <- rate_endpoints %>%
  filter(cause_name == "Diabetes mellitus", sex_name == "Both", measure_name == "DALYs (Disability-Adjusted Life Years)") %>%
  slice_max(relative_change_pct, n = 1, with_ties = FALSE)
overall_ratio_change <- yll_yld_ratios %>%
  filter(location_name == "Sub-Saharan Africa", year %in% c(1990, 2023)) %>%
  select(year, yll_to_yld_ratio) %>%
  pivot_wider(names_from = year, values_from = yll_to_yld_ratio, names_prefix = "ratio_")
```

# Data Readiness

```{r availability-table, results='asis'}
cat(format_pdf_table(
  availability %>%
    transmute(Input = input, Status = status, `Enabled analysis` = enabled_analysis),
  "Available inputs and enabled analyses",
  digits = 0,
  notes = c(
    "Age-standardized rates support comparisons across locations with different age structures.",
    "UN WPP population data can support demographic-only projections.",
    "Age-specific GBD rates are available and validated for GBD-consistent rate-effect decomposition and age-period-cohort analyses."
  )
))
```

The report presents descriptive, trend, forecast, sex-difference,
diabetes-type, and mortality-to-disability analyses for overall Sub-Saharan
Africa and its four GBD subregions. Matched GBD age-specific rates and counts
are now available for decomposition and age-period-cohort implementation.
Existing UN WPP population data support demographic projection scenarios.

\newpage

# Key Findings

For overall SSA, age-standardized diabetes prevalence increased from
`r round(overall_prevalence$value_1990, 1)` to
`r round(overall_prevalence$value_2023, 1)` per 100,000 between 1990 and 2023,
a `r round(overall_prevalence$relative_change_pct, 1)`% increase. Diabetes
mortality increased from `r round(overall_deaths$value_1990, 1)` to
`r round(overall_deaths$value_2023, 1)` per 100,000, a
`r round(overall_deaths$relative_change_pct, 1)`% increase.

Among the five regions, the highest 2023 age-standardized prevalence was in
`r clean_location(highest_2023_prevalence$location_name)` at
`r round(highest_2023_prevalence$value_2023, 1)` per 100,000. The fastest DALY
rate growth was in `r clean_location(fastest_daly_growth$location_name)`, with
a `r round(fastest_daly_growth$relative_change_pct, 1)`% increase from 1990 to
2023. The overall SSA YLL-to-YLD ratio changed from
`r round(overall_ratio_change$ratio_1990, 2)` in 1990 to
`r round(overall_ratio_change$ratio_2023, 2)` in 2023, indicating the evolving
balance between premature mortality and disability.

\newpage

# Measures and Interpretation

This analysis uses six GBD measures. They answer related but distinct
questions, so they should not be interpreted interchangeably.

```{r measure-definitions, results='asis'}
measure_definitions <- tibble(
  Measure = c("Prevalence", "Incidence", "Deaths", "YLLs", "YLDs", "DALYs"),
  `What it measures` = c(
    "Existing diabetes cases in a population at a given time.",
    "New diabetes cases occurring during a year.",
    "Deaths attributed to diabetes during a year.",
    "Years of life lost because diabetes deaths occurred before the reference life expectancy.",
    "Years lived with disability from diabetes and its sequelae.",
    "Total health loss: YLLs plus YLDs."
  ),
  `Main interpretation` = c(
    "How widespread diabetes is.",
    "How quickly new diabetes burden is being generated.",
    "Fatal burden.",
    "Premature-mortality burden.",
    "Non-fatal disability burden.",
    "Combined fatal and non-fatal burden."
  )
)
cat(format_pdf_table(
  measure_definitions,
  "GBD diabetes burden measures used in this analysis",
  digits = 0,
  max_rows = 6,
  notes = c(
    "Age-standardized rates are used for comparisons across regions.",
    "Counts measure absolute burden and are affected by population size and age structure."
  )
))
```

\newpage

#  Overall SSA Results

```{r main-burden-table}
main_measures <- c("Prevalence", "Incidence", "Deaths", "DALYs (Disability-Adjusted Life Years)")
main_burden_table <- rate_endpoints %>%
  filter(location_name == "Sub-Saharan Africa", measure_name %in% main_measures) %>%
  left_join(
    rate_trends %>%
      select(location_name, measure_name, cause_name, sex_name, annual_change_pct, lower_ci, upper_ci, p_value),
    by = c("location_name", "measure_name", "cause_name", "sex_name")
  ) %>%
  left_join(
    rate_forecasts %>%
      filter(year == 2030) %>%
      select(location_name, measure_name, cause_name, sex_name, selected_model, forecast, lower_95, upper_95),
    by = c("location_name", "measure_name", "cause_name", "sex_name")
  ) %>%
  filter(sex_name == "Both") %>%
  mutate(
    Measure = clean_measure(measure_name),
    Type = clean_cause(cause_name),
    `APC, % (95% CI)` = fmt_ci(annual_change_pct, lower_ci, upper_ci, digits = 2),
    `2030 forecast (95% PI)` = fmt_ci(forecast, lower_95, upper_95, digits = 1),
    `p` = fmt_p(p_value)
  ) %>%
  select(
    Measure, Type, `1990 rate` = value_1990, `2023 rate` = value_2023,
    `Change, %` = relative_change_pct, `APC, % (95% CI)`,
    Model = selected_model, `2030 forecast (95% PI)`, p
  )
write_csv(main_burden_table, file.path(table_dir, "overall_ssa_main_burden_table.csv"))
```

```{r main-burden-table-print, results='asis'}
cat(format_pdf_table(
  main_burden_table,
  "Age-standardized rates, trends, and forecasts for overall SSA",
  notes = c(
    "Rates are age-standardized GBD estimates per 100,000 population.",
    "APC is the average annual percentage change from a log-linear model.",
    "PI is the model-based prediction interval for the 2030 forecast.",
    "DALYs are disability-adjusted life years, combining fatal burden and non-fatal disability."
  )
))
```

```{r overall-burden-figure, fig.width=10.5, fig.height=6.8, fig.cap="Overall SSA diabetes burden by measure, diabetes type, and sex, 1990-2023."}
figure1_data <- rates %>%
  filter(location_name == "Sub-Saharan Africa", measure_name %in% main_measures) %>%
  left_join(measure_short_df, by = "measure_name") %>%
  mutate(
    measure = factor(measure, levels = c("Prevalence", "Incidence", "Deaths", "DALYs")),
    cause = factor(
      recode(cause_name, "Diabetes mellitus" = "DM",
             "Diabetes mellitus type 1" = "T1DM",
             "Diabetes mellitus type 2" = "T2DM"),
      levels = c("DM", "T1DM", "T2DM")
    )
  ) %>%
  mutate(
    facet_label = factor(
      paste(measure, cause, sep = ": "),
      levels = as.vector(outer(levels(measure), levels(cause), paste, sep = ": "))
    ),
    sex_name = factor(sex_name, levels = c("Male", "Female", "Both"))
  )

p_figure1 <- ggplot(figure1_data, aes(year, val, color = sex_name)) +
  geom_line(linewidth = 0.45) +
  facet_wrap(~ facet_label, scales = "free_y", ncol = 4) +
  scale_color_manual(values = c(Male = "#0072B2", Female = "#D55E00", Both = "#333333")) +
  scale_x_continuous(breaks = c(1990, 2000, 2010, 2020), expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(labels = rate_axis_labels) +
  labs(x = NULL, y = "Age-standardized rate per 100,000", color = NULL,
       caption = "Source: GBD 2023. Lines show point estimates.") +
  paper_theme(base_size = 8)
ggsave(file.path(figure_dir, "overall_ssa_burden_by_type_and_sex.png"), p_figure1, width = 10.5, height = 6.8, dpi = 300)
p_figure1
```

\newpage

# Overall SSA Extensions

```{r overall-rate-endpoints, results='asis'}
overall_rate_endpoints <- rate_endpoints %>%
  filter(location_name == "Sub-Saharan Africa", cause_name == "Diabetes mellitus", sex_name == "Both") %>%
  transmute(
    Measure = clean_measure(measure_name),
    `1990 rate (95% UI)` = fmt_ci(value_1990, lower_1990, upper_1990),
    `2023 rate (95% UI)` = fmt_ci(value_2023, lower_2023, upper_2023),
    `Change, %` = relative_change_pct
  )
cat(format_pdf_table(
  overall_rate_endpoints,
  "Overall SSA age-standardized endpoint rates",
  notes = c(
    "UI is the GBD 95 percent uncertainty interval.",
    "YLLs are years of life lost from premature mortality; YLDs are years lived with disability.",
    "DALYs are the sum of YLLs and YLDs."
  )
))
```

```{r forecast-figure, fig.width=10, fig.height=5.8, fig.cap="Overall SSA observed and forecast age-standardized rates through 2030."}
p_forecast <- plot_rate_forecasts(
  rates %>% filter(location_name == "Sub-Saharan Africa"),
  rate_forecasts %>% filter(location_name == "Sub-Saharan Africa"),
  "Sub-Saharan Africa"
) +
  labs(title = NULL, subtitle = NULL) +
  paper_theme(base_size = 8)
ggsave(file.path(figure_dir, "overall_ssa_age_standardized_rate_forecasts_2030.png"), p_forecast, width = 10, height = 5.8, dpi = 300)
p_forecast
```

```{r count-table, results='asis'}
overall_counts <- count_endpoints %>%
  filter(location_name == "Sub-Saharan Africa", cause_name == "Diabetes mellitus", sex_name == "Both") %>%
  transmute(
    Measure = clean_measure(measure_name),
    `1990 count (95% UI)` = fmt_ci(value_1990, lower_1990, upper_1990, digits = 0),
    `2023 count (95% UI)` = fmt_ci(value_2023, lower_2023, upper_2023, digits = 0),
    `Change, %` = relative_change_pct
  )
cat(format_pdf_table(
  overall_counts,
  "Overall SSA all-age count endpoints",
  notes = c(
    "Counts are all-age numbers, not age-standardized rates.",
    "YLLs and YLDs are time-equivalent health loss measures; DALYs equal YLLs plus YLDs."
  )
))
```

\newpage

# Overall Region Versus Subregions

```{r subregion-trends, fig.width=10, fig.height=5.8, fig.cap="Age-standardized total diabetes trends in overall SSA and four subregions."}
subregion_trend_data <- rates %>%
  filter(cause_name == "Diabetes mellitus", sex_name == "Both") %>%
  left_join(measure_short_df, by = "measure_name") %>%
  mutate(measure = factor(measure, levels = unname(measure_short)),
         region_short = factor(recode(location_name,
           "Sub-Saharan Africa" = "SSA",
           "Central Sub-Saharan Africa" = "Central SSA",
           "Eastern Sub-Saharan Africa" = "Eastern SSA",
           "Southern Sub-Saharan Africa" = "Southern SSA",
           "Western Sub-Saharan Africa" = "Western SSA"),levels = c("SSA", "Central SSA", "Eastern SSA", "Southern SSA", "Western SSA")))

new_region_palette <- c(
  "SSA" = "#111111",
  "Central SSA" = "#0072B2" ,
  "Eastern SSA" = "#009E73",
  "Southern SSA" = "#D55E00",
  "Western SSA" = "#CC79A7"
)
p_subregion_trends <- ggplot(subregion_trend_data, aes(year, val, color = region_short, linetype = region_short)) +
  geom_line(linewidth = 0.85) +
  facet_wrap(~ measure, scales = "free_y", ncol = 3) +
  scale_color_manual(values = new_region_palette) +
  scale_x_continuous(breaks = c(1990,1995, 2000,2005, 2010, 2015, 2020,2023), expand = expansion(mult = c(0.01, 0.05))) +
  scale_y_continuous(labels = rate_axis_labels) +
  labs(x = NULL, y = "Age-standardized rate per 100,000", color = NULL, linetype = NULL,
       caption = "Both sexes, total diabetes mellitus. Source: GBD 2023.") +
  paper_theme(base_size = 10) +
  theme(legend.text = element_text(size = 10, face="bold"),
        axis.text.x = element_text(size = 10, face="bold", angle=45, hjust=1),
        axis.text.y = element_text(size = 10, face="bold"))
ggsave(file.path(figure_dir, "ssa_subregion_age_standardized_rate_trends.png"), p_subregion_trends, width = 10, height = 5.8, dpi = 300)
p_subregion_trends
```

```{r subregion-endpoints, results='asis'}
subregion_endpoint_table <- rate_endpoints %>%
  filter(cause_name == "Diabetes mellitus", sex_name == "Both") %>%
  arrange(factor(location_name, levels = locations), factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    `1990 rate` = value_1990,
    `2023 rate` = value_2023,
    `Change, %` = relative_change_pct
  )
cat(format_pdf_table(
  subregion_endpoint_table,
  "Overall SSA and subregional age-standardized endpoint comparison",
  max_rows = 30,
  notes = c(
    "Rates are per 100,000 population and age-standardized.",
    "Change is the percentage change between 1990 and 2023."
  )
))
```

```{r subregion-trend-table, results='asis'}
subregion_trend_table <- rate_trends %>%
  filter(cause_name == "Diabetes mellitus", sex_name == "Both") %>%
  arrange(factor(location_name, levels = locations), factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    `APC, % (95% CI)` = fmt_ci(annual_change_pct, lower_ci, upper_ci, digits = 2),
    p = fmt_p(p_value),
    `Adj. R2` = adjusted_r_squared
  )
cat(format_pdf_table(
  subregion_trend_table,
  "Overall SSA and subregional full-period log-linear trends",
  digits = 3,
  max_rows = 30,
  notes = c(
    "APC is the average annual percentage change from a log-linear model across 1990-2023.",
    "Adj. R2 is the adjusted coefficient of determination for that trend model."
  )
))
```

```{r rate-ratio-to-ssa, results='asis'}
relative_to_ssa <- rates %>%
  filter(cause_name == "Diabetes mellitus", sex_name == "Both", year %in% c(1990, 2023)) %>%
  select(location_name, measure_name, year, val) %>%
  group_by(measure_name, year) %>%
  mutate(overall_ssa_rate = val[location_name == "Sub-Saharan Africa"]) %>%
  ungroup() %>%
  filter(location_name != "Sub-Saharan Africa") %>%
  mutate(subregion_to_overall_rate_ratio = val / overall_ssa_rate) %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    Year = year,
    `Rate ratio vs SSA` = subregion_to_overall_rate_ratio
  )
write_csv(relative_to_ssa, file.path(table_dir, "subregion_to_overall_ssa_rate_ratios.csv"))
cat(format_pdf_table(
  relative_to_ssa,
  "Subregional rate ratios relative to overall SSA",
  digits = 3,
  max_rows = 48,
  notes = c(
    "A ratio above 1 means the subregion has a higher age-standardized rate than overall SSA in that year.",
    "A ratio below 1 means the subregion has a lower rate than overall SSA."
  )
))
```

```{r percent-change-figure, fig.width=10, fig.height=4.8, fig.cap="Percent change in total diabetes age-standardized rates, 1990-2023."}
p_change <- subregion_endpoint_table %>%
  mutate(
    Measure = factor(Measure, levels = unname(measure_short)),
    Region = factor(Region, levels = clean_location(locations))
  ) %>%
  ggplot(aes(Region, `Change, %`, fill = Measure)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.65) +
  coord_flip() +
  scale_y_continuous(labels = label_percent(scale = 1),
                     limits =c(0,65)) +
  labs(x = NULL, y = "Percent change", fill = NULL,size = 10,
       caption = "Both sexes, total diabetes mellitus. Source: GBD 2023.") +
  paper_theme(base_size = 14)+
  theme(legend.text = element_text(size = 10, face="bold"),
        axis.text.x = element_text(size = 10, face="bold"),
        axis.text.y = element_text(size = 10, face="bold"))
ggsave(file.path(figure_dir, "ssa_subregion_percent_change_1990_2023.png"), p_change, width = 10, height = 4.8, dpi = 300)
p_change
```

\newpage

# Sex Differences and Diabetes Type Mix

```{r sex-ratio-table, results='asis'}
sex_ratio_table <- sex_ratios %>%
  filter(cause_name == "Diabetes mellitus", year == 2023) %>%
  arrange(factor(location_name, levels = locations), factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    Male,
    Female,
    `M:F ratio` = male_to_female_rate_ratio
  )
cat(format_pdf_table(
  sex_ratio_table,
  "Male-to-female age-standardized rate ratios in 2023",
  max_rows = 30,
  notes = "M:F ratio is the male age-standardized rate divided by the female age-standardized rate."
))
```

```{r type-share-table, results='asis'}
type_share_table <- type_shares %>%
  filter(year == 2023) %>%
  arrange(factor(location_name, levels = locations), factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    `T1DM share, %` = t1dm_share_pct,
    `T2DM share, %` = t2dm_share_pct,
    `Component difference, %` = component_sum_difference_pct
  )
cat(format_pdf_table(
  type_share_table,
  "Diabetes type shares of all-age burden in 2023",
  max_rows = 30,
  notes = c(
    "T1DM is type 1 diabetes mellitus; T2DM is type 2 diabetes mellitus.",
    "Component difference shows the percent difference between T1DM plus T2DM and total diabetes; small nonzero values reflect GBD estimation and rounding."
  )
))
```

\newpage

# Mortality-to-Disability Transition
```{r}
head(yll_yld_ratios)
## simplify the location names for plotting
yll_yld_ratios <- yll_yld_ratios %>%
  mutate(location_name = factor(recode(
    location_name,
    "Sub-Saharan Africa" = "SSA",
    "Central Sub-Saharan Africa" = "Central SSA",
    "Eastern Sub-Saharan Africa" = "Eastern SSA",
    "Southern Sub-Saharan Africa" = "Southern SSA",
    "Western Sub-Saharan Africa" = "Western SSA"
  ), levels = c("SSA", "Central SSA", "Eastern SSA", "Southern SSA", "Western SSA")))

```


```{r yll-yld-figure, fig.width=9.5, fig.height=4.5, fig.cap="YLL-to-YLD ratio in overall SSA and subregions, 1990-2023."}
p_ratio <- ggplot(yll_yld_ratios, aes(year, yll_to_yld_ratio, color = location_name)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5, na.rm = TRUE) +
  scale_color_manual(values = new_region_palette) +
  scale_x_continuous(breaks = c(1990,1995, 2000, 2005, 2010,2015, 2020, 2023)) +scale_y_continuous(breaks= seq((min(yll_yld_ratios$yll_to_yld_ratio)-0.5), (max(yll_yld_ratios$yll_to_yld_ratio)+1),length=20)) +
  labs(x = NULL, y = "YLL-to-YLD ratio", color = NULL,
       caption = "Both sexes, total diabetes mellitus. Ratio calculated from all-age counts.") +
  paper_theme(base_size = 8) +
  theme(legend.text = element_text(size = 10, face = "bold"),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"),
        axis.text.y = element_text(size = 10, face = "bold"))
ggsave(file.path(figure_dir, "ssa_subregion_yll_to_yld_ratio.png"), p_ratio, width = 9.5, height = 4.5, dpi = 300)
p_ratio
```

```{r yll-yld-period-table, results='asis'}
period_ratio_table <- period_ratios %>%
  transmute(Region = clean_location(location_name), Period = period, `YLL:YLD` = yll_to_yld_ratio)
cat(format_pdf_table(
  period_ratio_table,
  "Period mean YLL-to-YLD ratios",
  digits = 3,
  max_rows = 35,
  notes = "Higher YLL:YLD ratios indicate a larger share of burden from premature mortality relative to disability."
))
```

\newpage

# Pre-Forecast Diagnostics

Before fitting forecasting models, each overall SSA total-diabetes rate series
was inspected for smoothing behavior, autocorrelation structure, partial
autocorrelation structure, and differencing needs. These diagnostics are shown
for the six core GBD measures because they are the headline forecast outcomes.

```{r trailing-mean-figure, fig.width=10, fig.height=5.8, fig.cap="Observed rates and 5-year trailing means for overall SSA, total diabetes."}
p_trailing <- trailing_mean_data %>%
  left_join(measure_short_df, by = "measure_name") %>%
  mutate(measure = factor(measure, levels = unname(measure_short))) %>%
  ggplot(aes(year, val)) +
  geom_line(color = "#9E9E9E", linewidth = 0.65) +
  geom_line(aes(y = trailing_mean_5yr), color = "#0072B2", linewidth = 0.85, na.rm = TRUE) +
  facet_wrap(~ measure, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = c(1990,1995, 2000,2005, 2010,2015, 2020, 2023), expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(labels = rate_axis_labels) +
  labs(
    x = NULL, y = "Age-standardized rate per 100,000",
    caption = "Grey line: annual GBD estimate. Blue line: 5-year trailing mean."
  ) +
  paper_theme(base_size = 8)+
  theme(strip.text = element_text(face = "bold", size = 10),
        axis.text.x = element_text(size = 10, face="bold", angle=45, hjust=1),
        axis.text.y = element_text(size = 10, face="bold"))
ggsave(file.path(figure_dir, "overall_ssa_5yr_trailing_means.png"), p_trailing, width = 10, height = 5.8, dpi = 300)
p_trailing
```

```{r acf-pacf-figure, fig.width=10, fig.height=6.4, fig.cap="ACF and PACF diagnostics for overall SSA total diabetes rates."}
p_acf <- ggplot(acf_pacf_data, aes(lag, value)) +
  geom_hline(yintercept = 0, color = "#666666", linewidth = 0.25) +
  geom_segment(aes(xend = lag, yend = 0), color = "#0072B2", linewidth = 0.45) +
  facet_grid(diagnostic ~ measure) +
  scale_x_continuous(breaks = c(1, 4, 8, 12)) +
  labs(
    x = "Lag, years", y = "Correlation",
    caption = "ACF: autocorrelation function. PACF: partial autocorrelation function. Diagnostics shown for undifferenced annual rates."
  ) +
  paper_theme(base_size = 10) +
  theme(strip.text = element_text(face = "bold", size = 10),
        axis.text.x = element_text(size = 10, face="bold"),
        axis.text.y = element_text(size = 10, face="bold"))
ggsave(file.path(figure_dir, "overall_ssa_acf_pacf_diagnostics.png"), p_acf, width = 10, height = 6.4, dpi = 300)
p_acf
```

```{r stationarity-table, results='asis'}
cat(format_pdf_table(
  stationarity_checks %>%
    transmute(
      Measure,
      `KPSS ndiffs` = kpss_ndiffs,
      `ADF ndiffs` = adf_ndiffs,
      `PP ndiffs` = pp_ndiffs,
      `ARIMA p` = auto_arima_p,
      `ARIMA d` = auto_arima_d,
      `ARIMA q` = auto_arima_q,
      `Selected ARIMA` = selected_arima
    ),
  "Stationarity and ARIMA differencing diagnostics for overall SSA total diabetes rates",
  digits = 0,
  max_rows = 6,
  notes = c(
    "ndiffs is the number of differences suggested by each unit-root or stationarity test.",
    "KPSS tests stationarity as the null hypothesis; ADF and PP test a unit root as the null hypothesis.",
    "The selected ARIMA order is the order returned by auto.arima for the undifferenced input series."
  )
))
```

In `auto.arima()`, the differencing order \(d\) is selected before searching
over AR and MA orders. For a non-seasonal annual series, the function uses
unit-root and stationarity tests through `ndiffs()` to estimate the smallest
number of differences needed to make the series approximately stationary. In
this report, KPSS, ADF, and PP suggestions are shown side by side, while the
`auto_arima_d` column records the value actually used by `auto.arima()` in the
selected model. After \(d\) is chosen, `auto.arima()` searches over candidate
\((p,d,q)\) models and chooses the specification with the best information
criterion, subject to its default constraints.

\newpage

# Forecasting Models and Validation

The forecasting target is the annual age-standardized rate \(y_t\) for a
single stratum defined by location, measure, diabetes cause, and sex. Calendar
year \(t\) is the time index. Models are trained on observed GBD estimates from
1990 to 2023 and evaluated using expanding-window one-step-ahead validation.
The initial training window ends in 2004; the model is then refit repeatedly to
forecast each next year through 2023.

```{r model-registry-table, results='asis'}
cat(format_pdf_table(
  model_registry %>%
    transmute(Model = model, Status = status, `Role in analysis` = primary_use),
  "Forecasting model registry",
  digits = 0,
  max_rows = 4,
  notes = c(
    "LSTM fitting requires the R keras/tensorflow packages plus a working Python TensorFlow backend.",
    lstm_status$lstm_status
  )
))
```

## Model Specifications

**Log-linear trend model.** The descriptive trend model is
\[
\log(y_t) = \alpha + \beta t + \epsilon_t.
\]
The average annual percentage change is
\[
APC = 100\{\exp(\beta)-1\}.
\]
This model is used for interpretable long-run change, not for selecting the
best forecast.

**ARIMA.** The autoregressive integrated moving-average model is written as
\[
\phi(B)(1-B)^d y_t = c + \theta(B)\epsilon_t,
\]
where \(B\) is the backshift operator, \(d\) is the differencing order, and
\(\phi(B)\) and \(\theta(B)\) are autoregressive and moving-average
polynomials. ARIMA is included because diabetes rates may contain serial
dependence, gradual trend, and short-term autocorrelated deviations.

**ETS.** The exponential-smoothing state-space model decomposes the series into
level, trend, and error components. In the additive-error form,
\[
y_t = \ell_{t-1} + b_{t-1} + \epsilon_t,
\]
\[
\ell_t = \ell_{t-1} + b_{t-1} + \alpha\epsilon_t,\quad
b_t = b_{t-1} + \beta\epsilon_t.
\]
ETS is included because it performs well for smooth epidemiologic trajectories
where recent observations should receive more weight than distant observations.

**LSTM.** The long short-term memory model is specified as a recurrent neural
network for lagged first-differenced rate sequences,
\(\Delta y_t = y_t - y_{t-1}\). The fitted network predicts
\(\widehat{\Delta y}_{t+1}\), and the rate forecast is reconstructed as
\(\hat{y}_{t+1} = y_t + \widehat{\Delta y}_{t+1}\). For a scaled input vector
\(x_t\), the core gates are
\[
f_t = \sigma(W_f x_t + U_f h_{t-1} + b_f),
\]
\[
i_t = \sigma(W_i x_t + U_i h_{t-1} + b_i),\quad
\tilde{c}_t = \tanh(W_c x_t + U_c h_{t-1} + b_c),
\]
\[
c_t = f_t \odot c_{t-1} + i_t \odot \tilde{c}_t,\quad
o_t = \sigma(W_o x_t + U_o h_{t-1} + b_o),
\]
\[
h_t = o_t \odot \tanh(c_t).
\]
The forecast is generated from the hidden state \(h_t\). LSTMs are theoretically
attractive because the gated cell state can represent nonlinear temporal
patterns and longer memory than standard autoregressive models. They are not
assumed to outperform ARIMA automatically; performance must be judged by the
validation metrics below.

## Variables Used

All forecasting models use the same outcome: annual age-standardized rate per
100,000 population. Models are fit separately for each location, measure,
diabetes cause, and sex. No external covariates are included in the current
forecasting stage; therefore, forecasts represent continuation of historical
rate patterns, not causal projections under policy or exposure scenarios.

## Model Comparison Metrics

For validation year \(t\), let \(y_t\) be the observed rate and \(\hat{y}_t\)
the one-step-ahead forecast. With \(n\) validation years:
\[
RMSE = \sqrt{\frac{1}{n}\sum_{t=1}^{n}(y_t-\hat{y}_t)^2},
\]
\[
MAE = \frac{1}{n}\sum_{t=1}^{n}|y_t-\hat{y}_t|,
\]
\[
MAPE = \frac{100}{n}\sum_{t=1}^{n}\left|\frac{y_t-\hat{y}_t}{y_t}\right|,
\]
\[
MASE =
\frac{\frac{1}{n}\sum_{t=1}^{n}|y_t-\hat{y}_t|}
{\frac{1}{T-1}\sum_{t=2}^{T}|y_t-y_{t-1}|}.
\]
Lower values indicate better validation performance. MASE is the primary
selection metric because it is scale-free and compares model error with a
naive random-walk benchmark.

For LSTM forecasts, 95 percent prediction intervals are estimated from the
empirical distribution of expanding-window validation residuals. The interval
is centered on the recursive LSTM point forecast and widened by the square
root of the forecast horizon. These intervals are therefore validation-residual
intervals rather than parametric state-space intervals.

## Model Comparison Results

The production 2030 forecast is selected from ARIMA and ETS for every
location-measure-cause-sex stratum because those models provide stable
prediction intervals for all series. The headline model comparison separately
fits ARIMA, ETS, and LSTM for overall SSA, both sexes, total diabetes across
all six measures. These headline results are used to compare machine-learning
validation performance against classical time-series models.

```{r headline-model-comparison-table, results='asis'}
# Clean the headline comparison so model metrics are readable in the PDF.
headline_model_accuracy_table <- headline_model_accuracy %>%
  mutate(
    Measure = clean_measure(measure_name),
    Model = model,
    Status = if_else(is.na(mase), "Not fitted", "Fitted")
  ) %>%
  arrange(factor(Measure, levels = unname(measure_short)), Model) %>%
  transmute(
    Measure,
    Model,
    RMSE = rmse,
    MAE = mae,
    MAPE = mape,
    MASE = mase,
    `Validation years` = validation_years,
    Selected = if_else(selected, "Yes", "No"),
    Status
  )

cat(format_pdf_table(
  headline_model_accuracy_table,
  "Headline forecast model comparison for overall SSA total diabetes rates",
  digits = 3,
  max_rows = 18,
  notes = c(
    "The headline comparison uses expanding-window one-step-ahead validation.",
    "LSTM is fitted to lagged first-differenced rates and transformed back to the original rate scale for validation.",
    "The selected model is the one with the lowest MASE within each measure."
  )
))
```

```{r headline-selected-models-table, results='asis'}
# Show the winning headline model for each measure so LSTM selections are explicit.
headline_selected_models <- headline_model_accuracy %>%
  filter(selected) %>%
  arrange(factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Measure = clean_measure(measure_name),
    `Selected headline model` = model,
    RMSE = rmse,
    MAE = mae,
    MAPE = mape,
    MASE = mase
  )

write_csv(headline_selected_models, file.path(table_dir, "headline_selected_forecast_models.csv"))

cat(format_pdf_table(
  headline_selected_models,
  "Selected headline forecast model by measure",
  digits = 3,
  max_rows = 6,
  notes = c(
    "Selection is based on the lowest MASE among ARIMA, ETS, and LSTM.",
    "This table applies to overall SSA, both sexes, total diabetes."
  )
))
```

```{r model-selection-summary-table, results='asis'}
# Count how often each production model is selected across all forecast strata.
model_selection_summary <- forecast_accuracy %>%
  filter(selected) %>%
  count(model, name = "Selected strata") %>%
  mutate(`Share selected, %` = 100 * `Selected strata` / sum(`Selected strata`)) %>%
  arrange(desc(`Selected strata`)) %>%
  rename(Model = model)

write_csv(model_selection_summary, file.path(table_dir, "forecast_model_selection_summary.csv"))

cat(format_pdf_table(
  model_selection_summary,
  "Production forecast model selection across all strata",
  digits = 1,
  max_rows = 4,
  notes = c(
    "A stratum is one unique location, measure, diabetes cause, and sex combination.",
    "This production table compares ARIMA and ETS for all strata; LSTM results are shown in the headline comparison above."
  )
))
```

```{r headline-model-mase-figure, fig.width=9.5, fig.height=4.8, fig.cap="Headline forecast model comparison by MASE."}
# Plot MASE because it is the scale-free model-selection metric.
p_model_mase <- headline_model_accuracy %>%
  filter(!is.na(mase)) %>%
  mutate(
    measure = factor(clean_measure(measure_name), levels = unname(measure_short)),
    model = factor(model, levels = c("ARIMA", "ETS", "LSTM"))
  ) %>%
  ggplot(aes(model, mase, fill = model)) +
  geom_col(width = 0.65) +
  facet_wrap(~ measure, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c(ARIMA = "#0072B2", ETS = "#D55E00", LSTM = "#009E73")) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    x = NULL, y = "MASE",
    caption = "Lower MASE indicates better one-step-ahead validation performance."
  ) +
  paper_theme(base_size = 8)
ggsave(file.path(figure_dir, "headline_forecast_model_mase_comparison.png"), p_model_mase, width = 9.5, height = 4.8, dpi = 300)
p_model_mase
```

```{r headline-validation-predictions-figure, fig.width=10, fig.height=5.8, fig.cap="Observed and predicted validation-year rates for headline forecast models."}
# Compare one-step-ahead validation predictions against observed rates.
p_validation_predictions <- headline_model_errors %>%
  left_join(measure_short_df, by = "measure_name") %>%
  mutate(
    measure = factor(measure, levels = unname(measure_short)),
    model = factor(model, levels = c("ARIMA", "ETS", "LSTM"))
  ) %>%
  ggplot(aes(validation_year)) +
  geom_line(aes(y = actual), color = "#111111", linewidth = 0.45) +
  geom_point(aes(y = actual), color = "#111111", size = 0.9) +
  geom_line(aes(y = predicted, color = model), linewidth = 0.5) +
  facet_wrap(~ measure, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(ARIMA = "#0072B2", ETS = "#D55E00", LSTM = "#009E73")) +
  scale_x_continuous(breaks = c(2015, 2018, 2020, 2023), expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(labels = rate_axis_labels) +
  labs(
    x = NULL, y = "Age-standardized rate per 100,000", color = NULL,
    caption = "Black line: observed GBD rate. Colored lines: one-step-ahead validation predictions."
  ) +
  paper_theme(base_size = 8)
ggsave(file.path(figure_dir, "headline_forecast_validation_predictions.png"), p_validation_predictions, width = 10, height = 5.8, dpi = 300)
p_validation_predictions
```

# Forecast Validation and 2030 Projections

```{r headline-selected-forecast-table, results='asis'}
# Show 2030 forecasts from the selected ARIMA/ETS/LSTM headline model.
headline_selected_forecast_2030 <- headline_selected_forecasts %>%
  filter(year == 2030) %>%
  arrange(factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Measure = clean_measure(measure_name),
    Model = selected_model,
    `2030 forecast (95% PI)` = fmt_ci(forecast, lower_95, upper_95),
    `Interval method` = interval_method
  )

cat(format_pdf_table(
  headline_selected_forecast_2030,
  "Headline selected-model 2030 forecasts with prediction intervals",
  digits = 1,
  max_rows = 6,
  notes = c(
    "This table applies to overall SSA, both sexes, total diabetes.",
    "ARIMA and ETS intervals are model prediction intervals.",
    "LSTM intervals are empirical validation-residual prediction intervals."
  )
))
```

```{r headline-selected-forecast-figure, fig.width=10, fig.height=5.8, fig.cap="Headline selected-model forecasts through 2030 with prediction intervals."}
# Plot selected headline forecasts, including LSTM residual intervals where selected.
headline_observed <- rates %>%
  filter(
    location_name == "Sub-Saharan Africa",
    cause_name == "Diabetes mellitus",
    sex_name == "Both"
  ) %>%
  transmute(
    measure_name,
    year,
    value = val,
    lower = NA_real_,
    upper = NA_real_,
    selected_model = "Observed",
    series = "Observed"
  )

headline_projected <- headline_selected_forecasts %>%
  transmute(
    measure_name,
    year,
    value = forecast,
    lower = lower_95,
    upper = upper_95,
    selected_model,
    series = "Forecast"
  )

p_headline_selected_forecast <- bind_rows(headline_observed, headline_projected) %>%
  left_join(measure_short_df, by = "measure_name") %>%
  mutate(
    measure = factor(measure, levels = unname(measure_short)),
    selected_model = factor(selected_model, levels = c("Observed", "ARIMA", "ETS", "LSTM"))
  ) %>%
  ggplot(aes(year, value, color = selected_model)) +
  geom_ribbon(
    data = ~ filter(.x, series == "Forecast"),
    aes(ymin = lower, ymax = upper, fill = selected_model),
    alpha = 0.18, color = NA
  ) +
  geom_line(linewidth = 0.55) +
  geom_vline(xintercept = 2023.5, linewidth = 0.3, linetype = "dashed") +
  facet_wrap(~ measure, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(Observed = "#111111", ARIMA = "#0072B2", ETS = "#D55E00", LSTM = "#009E73")) +
  scale_fill_manual(values = c(Observed = "#111111", ARIMA = "#0072B2", ETS = "#D55E00", LSTM = "#009E73")) +
  scale_x_continuous(breaks = c(1990, 2000, 2010, 2020, 2030), expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(labels = rate_axis_labels) +
  labs(
    x = NULL, y = "Age-standardized rate per 100,000",
    color = NULL, fill = NULL,
    caption = "Forecast ribbons show 95% prediction intervals. LSTM intervals use validation residuals."
  ) +
  paper_theme(base_size = 8)

ggsave(
  file.path(figure_dir, "headline_selected_model_forecasts_2030.png"),
  p_headline_selected_forecast, width = 10, height = 5.8, dpi = 300
)
p_headline_selected_forecast
```

```{r subregion-forecast-interval-table, results='asis'}
# Show 2030 subregion prediction intervals from the full-strata production forecasts.
subregion_forecast_2030 <- rate_forecasts %>%
  filter(
    cause_name == "Diabetes mellitus",
    sex_name == "Both",
    year == 2030
  ) %>%
  arrange(factor(location_name, levels = locations), factor(measure_name, levels = measure_levels)) %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    Model = selected_model,
    `2030 forecast (95% PI)` = fmt_ci(forecast, lower_95, upper_95)
  )

write_csv(subregion_forecast_2030, file.path(table_dir, "subregion_2030_prediction_intervals.csv"))

cat(format_pdf_table(
  subregion_forecast_2030,
  "Overall SSA and subregional 2030 prediction intervals",
  digits = 1,
  max_rows = 30,
  notes = c(
    "Forecasts are age-standardized rates per 100,000 for both sexes and total diabetes.",
    "Subregion forecasts use the full-strata ARIMA/ETS production models.",
    "The LSTM-inclusive headline forecast for overall SSA is shown separately above."
  )
))
```

The LSTM-inclusive headline forecasts above are the main forecast results for
overall SSA, both sexes, total diabetes. Full-strata ARIMA/ETS production
outputs are retained as machine-readable supplementary files because they cover
every location, sex, measure, and diabetes type. The full production ARIMA and
ETS model comparison is saved as
`outputs/sub_saharan_africa/full_analysis/tables/forecast_model_accuracy.csv`.
The headline ARIMA/ETS/LSTM comparison is saved as
`outputs/sub_saharan_africa/full_analysis/tables/headline_forecast_model_comparison.csv`.
The headline selected-model forecasts with LSTM prediction intervals are saved
as `outputs/sub_saharan_africa/full_analysis/tables/headline_selected_model_forecasts_2024_2030.csv`.

\newpage

# Age-Specific Count Quality

```{r age-count-qc-table, results='asis'}
cat(format_pdf_table(
  age_count_qc %>%
    transmute(Region = clean_location(location_name), Sex = sex_name, Rows = rows, Measures = measures, Causes = causes, Ages = ages, Years = years),
  "Age-specific count quality summary",
  digits = 0,
  notes = "Each complete region-sex stratum should contain 34 years, 25 age groups, 6 measures, and 3 diabetes causes."
))
```

Age-specific counts, matched GBD age-specific rates, and UN WPP population data
are complete. The rate file covers all five regions, three sexes, three causes,
six measures, 25 mutually exclusive age groups, and every year from 1990 to
2023.

\newpage

# Symmetric Decomposition of Change, 1990-2023

```{r population-validation-table, results='asis'}
cat(format_pdf_table(
  population_qc %>%
    transmute(
      `Demographic cells` = demographic_cells,
      `Cells reconstructed` = cells_with_denominator,
      `Maximum relative spread` = maximum_relative_spread,
      `Median relative spread` = median_relative_spread
    ),
  "Validation of GBD-implied age-sex population denominators",
  digits = 8,
  notes = "Population was reconstructed as count/rate x 100,000. Agreement across causes and measures confirms that matched counts and rates use a common denominator."
))
```

```{r decomposition-table, results='asis'}
decomposition_headline <- symmetric_decomposition %>%
  filter(sex_name == "Both", cause_name == "Diabetes mellitus") %>%
  transmute(
    Region = clean_location(location_name),
    Measure = clean_measure(measure_name),
    `Observed change` = observed_change,
    `Population size` = population_size_effect,
    `Age structure` = age_structure_effect,
    `Age-specific rate` = age_specific_rate_effect,
    `Numerical error` = decomposition_error
  ) %>%
  arrange(factor(Region, levels = c("SSA", "Central SSA", "Eastern SSA", "Southern SSA", "Western SSA")),
          factor(Measure, levels = unname(measure_short)))
cat(format_pdf_table(
  decomposition_headline,
  "Symmetric decomposition of the change in total diabetes burden, both sexes, 1990-2023",
  digits = 1,
  max_rows = 30,
  notes = c(
    "Effects are absolute changes in GBD counts and sum to the reconstructed observed change.",
    "Each effect is the average marginal contribution across all six possible factor orderings."
  )
))
```

```{r decomposition-figure, fig.width=7.2, fig.height=4.8, fig.cap="Symmetric decomposition of the change in total diabetes burden, 1990-2023."}
decomposition_plot_data <- symmetric_decomposition %>%
  filter(
    sex_name == "Both", cause_name == "Diabetes mellitus",
    measure_name %in% c("Prevalence", "Deaths", "DALYs (Disability-Adjusted Life Years)")
  ) %>%
  select(location_name, measure_name, population_size_effect,
         age_structure_effect, age_specific_rate_effect) %>%
  pivot_longer(
    ends_with("effect"), names_to = "effect", values_to = "contribution"
  ) %>%
  mutate(
    effect = recode(effect,
      population_size_effect = "Population size",
      age_structure_effect = "Age structure",
      age_specific_rate_effect = "Age-specific rate"),
    location_name = factor(location_name, levels = locations),
    measure = factor(clean_measure(measure_name), levels = c("Prevalence", "Deaths", "DALYs"))
  )
decomposition_plot <- ggplot(decomposition_plot_data,
  aes(x = location_name, y = contribution, fill = effect)) +
  geom_col(position = "stack") +
  geom_hline(yintercept = 0, linewidth = 0.25) +
  facet_wrap(~ measure, scales = "free_y", ncol = 1) +
  scale_x_discrete(labels = clean_location) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("Population size" = "#0072B2", "Age structure" = "#D55E00", "Age-specific rate" = "#009E73")) +
  labs(x = NULL, y = "Contribution to count change") + paper_theme() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
print(decomposition_plot)
ggsave(file.path(figure_dir, "symmetric_decomposition_1990_2023.png"),
       decomposition_plot, width = 7.2, height = 4.8, dpi = 300)
```

\newpage

# Age-Specific Patterns and Age-Adjusted Temporal Drift

```{r age-pattern-figure, fig.width=7.2, fig.height=4.8, fig.cap="Age-specific total diabetes rates in 2023, both sexes."}
age_pattern_plot_data <- age_specific_endpoints %>%
  filter(
    sex_name == "Both", cause_name == "Diabetes mellitus",
    measure_name %in% c("Prevalence", "Incidence", "Deaths", "DALYs (Disability-Adjusted Life Years)")
  ) %>%
  mutate(
    age_name = factor(age_name, levels = c(
      "0-6 days", "7-27 days", "1-5 months", "6-11 months", "12-23 months",
      "2-4 years", "5-9 years", "10-14 years", "15-19 years", "20-24 years",
      "25-29 years", "30-34 years", "35-39 years", "40-44 years", "45-49 years",
      "50-54 years", "55-59 years", "60-64 years", "65-69 years", "70-74 years",
      "75-79 years", "80-84 years", "85-89 years", "90-94 years", "95+ years")),
    measure = factor(clean_measure(measure_name), levels = c("Prevalence", "Incidence", "Deaths", "DALYs")),
    location_name = factor(location_name, levels = locations)
  )
age_pattern_plot <- ggplot(age_pattern_plot_data,
  aes(x = age_name, y = rate_2023, color = location_name, group = location_name)) +
  geom_line(linewidth = 0.45) + geom_point(size = 0.65) +
  facet_wrap(~ measure, scales = "free_y", ncol = 2) +
  scale_color_manual(values = region_palette, labels = clean_location) +
  scale_y_continuous(labels = comma) +
  labs(x = "Age group", y = "Rate per 100,000") + paper_theme() +
  theme(
    axis.text.x = element_text(angle = 65, hjust = 1, size = 5),
    legend.text = element_text(size = 7)
  )
print(age_pattern_plot)
ggsave(file.path(figure_dir, "age_specific_rates_2023.png"),
       age_pattern_plot, width = 7.2, height = 4.8, dpi = 300)
```

```{r drift-table, results='asis'}
cat(format_pdf_table(
  age_adjusted_drift %>%
    filter(sex_name == "Both", cause_name == "Diabetes mellitus") %>%
    transmute(
      Region = clean_location(location_name), Measure = measure_name,
      `Annual drift (%)` = annual_drift_pct,
      `Lower 95%` = lower_95, `Upper 95%` = upper_95, `P value` = p_value
    ),
  "Age-adjusted annual drift in total diabetes incidence and mortality, 1990-2023",
  digits = 3,
  notes = c(
    "Quasi-Poisson log-linear models include categorical age effects and GBD-implied population offsets.",
    "These estimates quantify age-adjusted temporal drift; they are not presented as a fully identified age-period-cohort decomposition."
  )
))
```

\newpage

# Population Reconstruction

```{r reconstructed-population-table, results='asis'}
reconstructed_population_summary <- age_population %>%
  filter(year %in% c(1990, 2023), sex_name == "Both") %>%
  group_by(location_name, year) %>%
  summarise(population = sum(population), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = population, names_prefix = "Population ") %>%
  mutate(`Change (%)` = 100 * (`Population 2023` / `Population 1990` - 1)) %>%
  transmute(Region = clean_location(location_name), across(everything()))
cat(format_pdf_table(
  reconstructed_population_summary,
  "GBD-implied population reconstructed from matched age-specific counts and rates",
  digits = 0,
  notes = "Both-sex population totals are sums of the 25 mutually exclusive age groups."
))
```

\newpage

# Supplementary Output Catalogue

```{r catalogue, results='asis'}
catalogue <- tibble(
  file = basename(list.files(table_dir, full.names = TRUE)),
  directory = table_dir
) %>%
  arrange(file)
cat(format_pdf_table(
  catalogue,
  "Machine-readable table catalogue",
  digits = 0,
  max_rows = 30,
  notes = "CSV tables contain the full machine-readable outputs used to create the compact report tables."
))
```

\newpage

# Methods Notes

This report uses GBD 2023 point estimates and uncertainty intervals for
Sub-Saharan Africa and the four GBD subregions. Age-standardized rates are
reported per 100,000 population. Full-period APC estimates come from
log-linear models and are descriptive summaries of average annual change.
Forecast intervals reflect model uncertainty only; GBD uncertainty in
forecasts should be propagated later if draws become available. The symmetric
decomposition averages effects across all factor orderings. Age-adjusted drift
models use quasi-Poisson regression with categorical age effects and should not
be interpreted as a fully identified age-period-cohort decomposition.
